How environmental drivers of spatial synchrony interact

### 2023-02-08 Paper/Paper.Rnw
minor changes suggested by coauthors, planning to submit this version to Ecography

@@

\affil[a]{Department of Ecology and Evolutionary Biology and Center for Ecological Research, University of Kansas}
\affil[b]{Department of Environmental Sciences, University of Virginia}
\affil[c]{Department of Geography, University of California, Los Angeles}
\affil[d]{Marine Biological Association of the United Kingdom}
\affil[e]{Center for Watershed Sciences, University of California, Davis}
\affil[f]{Applied Ocean Physics and Engineering Department, Woods Hole Oceanographic Institution}
\affil[g]{Earth Research Institute, University of California, Santa Barbara}
\affil[*]{Corresponding author, reuman@ku.edu}

@@

%Intro to Moran effects and history of the early years where people had trouble identifying them
Correlations between weather time series measured in different locations can 
induce synchrony between populations in those locations if the weather variables influence
population processes. This mechanism, called the Moran effect, is now known to be one of the most
important causes of synchrony. But the mechanism was originally proposed 
theoretically \citep{Moran1953}, and decades passed during which it was 
considered difficult to distinguish this potential cause of synchrony from others in field systems.
Synchrony has long been thought to have three causes: dispersal, and trophic interactions with a synchronous or
mobile species, in addition to the Moran effect \citep{Liebhold2004}. But using common past statistical approaches
which focussed on declines in population correlations with distance, patterns of synchrony produced
by each of these mechanisms are quite similar \citep{Ranta1999,Abbott2007,Liebhold2004,Walter2017}; so examination
of such patterns provides little or no traction for inferring the causes of synchrony. 
Early papers demonstrating the Moran effect mechanism resorted to special cases where dispersal was 
impossible and predators were absent \citep{Grenfell1998,Tedesco2004}. Controlled experiments also confirmed that all
three theorized mechanisms could be involved in synchrony [e.g., \citealp{Vasseur2009}]. But the problem of inferring
the relative contribution of the three mechanisms in field systems was considered a challenge 
until recently.

%Recent statistical explorations of Moran effects
Recent research has provided new statistical viewpoints which have, when sufficient data are available, 
essentially solved the problem of inference
of the causes of synchrony, and the research has revealed the broad importance of Moran effects. 
%geographic approaches
Approaches based on formal statistical comparisons between population synchrony maps and maps of 
environmental correlations produced inferences that precipitation and temperature Moran effects are important
causes of synchrony of terrestrial \citep{Defriez2017b} and marine \citep{Defriez2017} primary productivity. 
Another geographic approach, based on multiple regression of distance matrices (MRM), was used in an early paper
to infer that a precipitation Moran effect is a cause of synchrony for the spongy moth \citep{Haynes2013},
an invasive forest pest whose name was recently changed from ``gypsy moth.'' 
Geographic approaches to identifying causes of synchrony, often MRM approaches 
specifically \citep{Koenig2017,Walter2020,Bogdziewicz2021}, 
have become widespread, and the geography of synchrony has become a mainstream area of study \citep{Walter2017}.
The approach of \cite{Gouveia2016}, called ``noncentered local indicators of spatial association'' (ncLISA), is another 
geographic approach; as is the ``fluvial synchrogram'' of \cite{Larsen2021}, a tool for studying synchrony in 
river networks.
MRM approaches have also been used to identify dispersal as a cause of synchrony \citep{Anderson2018},
sometimes combined with genetic methods \cite{Haynes2022}.
%Time and timescale methods
Another class of methods exploit the time and timescale structure of synchrony.
Using a method that examined the match between temporal variations in population and environmental synchrony, 
\cite{Allstadt2015} confirmed the importance of a precipitation Moran effect for synchrony of spongy moth.
A suite of wavelet and related Fourier techniques have been developed that can comprehensively describe the time and timescale
structure of synchrony, identify Moran drivers of synchrony opperating in distinct timescale bands, and apportion fractions
of observed synchrony to respective Moran drivers \citep{Sheppard2016,Sheppard2017,Desharnais2018,Sheppard2019,wsyn,Anderson2021}.
The methods have been used, for instance, to provide one of the first clear demonstrations that changes in 
Moran effects are another of the impacts of climatic changes \citep{Sheppard2016,Sheppard2019}; and to illustrate the importance of the
timescale structure of environmental covariation for altering Moran effects \citep{Desharnais2018}.
These methods have since been used to identify Moran effects in several additional systems
(e.g., \citealp{Anderson2019,Walter2019,Walter2019,GarciaCarreras2022,CastoraniSheppardKelp}).

%Intro to interactions
The techniques reviewed above have made it possible for several recent papers to identify cases in which two or more
distinct Moran drivers opperate simultaneously on the same populations 
(e.g., \citealp{Defriez2017b,Haynes2019,Walter2019,Walter2020,Anderson2021}).
In fact, two recent papers have documented that Moran effects of distinct 
environmental drivers can interact, either synergistically or antagonistically,
so that total population synchrony can be either greater than or less than what synchrony would be if the distinct Moran 
drivers operated independently \citep{Sheppard2019,CastoraniSheppardKelp}. 
\cite{Sheppard2019} showed that long-timescale ($>4$y period) synchrony in a chlorophyll density index 
in the seas around the United Kingdom is substantially augmented by interactions between environmental 
drivers of synchrony and drivers linked with consumption by a copepod consumer.
\cite{CastoraniSheppardKelp} showed that nutrient dynamics and wave disturbance, two Moran drivers
of synchrony in giant kelp populations on the California (CA) coast, interact either synergistically 
or destructively, depending on which portion of the CA coast is examined and on 
the timescale of analysis. 
The main purpose of this study is to establish a general theory of the mechanism
of interacting Moran effects and to use examples and applications of the theory to provide 
ecological intuition for how the mechanism of interacting Moran effects works. Our goals are
distinct from those of \cite{Kendall2000}, who examined interactions between Moran and dispersal
causes of synchrony.

@@

%Direct Moran effects
Direct Moran effects for kelp and how they manifest in our theory are fairly straightforward. 
Nitrates are henceforth identified with $\epsilon^{(1)}$ and waves with $\epsilon^{(2)}$. 
Both nitrates and waves fluctuate seasonally in CA \citep{Schiel2015}. 
Thus nitrate and wave 
synchrony had a strong annual component (Fig. \ref{fig:noise}a,b,j,k for the CCA1 and SB regions,
Fig. \ref{SM-fig:directMoran_SM_lag4} for CCA2), which produced 
some of the annual synchrony observed in kelp 
(Fig. \ref{fig:SummaryKelpFig}a,d, green and blue lines).
Nitrates and waves are also synchronous on long timescales $>8$y  
(Fig. \ref{fig:noise}g,h,p,q for the CCA1 and SB regions, Fig. \ref{SM-fig:directMoran_SM_lag4} 
for CCA2), possibly due to the North Pacific Gyre Oscillation 
\citep{CastoraniSheppardKelp, DiLorenzo2008}. The long-timescale synchrony
in nitrates and waves produced some of the long-timescale
synchrony in kelp (Fig. \ref{fig:SummaryKelpFig}c,f, green and blue lines).
Kelp synchrony was stronger in CCA1 than in SB 
(Fig. \ref{fig:SummaryKelpFig}a-c v. d-f, black lines) in part because the synchrony of nitrates, and of 
waves, was more pronounced in CCA1 than in SB 
(Fig. \ref{fig:noise}), and also because waves
had a stronger influence on kelp in central California than in SB:
regression coefficients determining $f_{\mcP^{(1)}}$, $f_{\mcP^{(2)}}$
and $f_{\mcB}$ are in Table \ref{SM-tab:regcoefs}.
%***DAN: Later you will have to make sure all rules are followed for citing all sup mat figures and 
%tables and sections from the main text, in order. I think currently they are not all cited, in part
%because I have been shifting around what information is where in the whole paper.

@@

%*DONE Climate change
Climate change has the potential to influence interactions between Moran effects in two specific ways
which can be illuminated by our theory, and this potential should be investigated in
future work. Examining the third term of (\ref{rhoww_T1to3}), climate change could
alter interaction effects if it: 1) alters the term $\rho_{\epsilon^{(1)} \epsilon^{(2)}}$
quantifying cross synchrony between Moran drivers; or 2) alters one of the terms
$f_{\mcP^{(1)}}$ or $f_{\mcP^{(2)}}$ specifying the nature of the influence of the 
environmental variables $\epsilon^{(1)}$ and $\epsilon^{(2)}$, respectively, on populations. As advocated above,
the term $\rho_{\epsilon^{(1)} \epsilon^{(2)}}$ should be systematically computed in future work, for a variety 
of environmental variables, to assess whether interactions between Moran effects are likely to
be general. As part of that process, the potential for changes in $\rho_{\epsilon^{(1)} \epsilon^{(2)}}$
could also be assessed, by using either time-windowed versions or wavelet adaptations 
of this statistic, applied to either long-term climate records or future climate scanarios. 
Differences in $f_{\mcP^{(1)}}$ were responsible for differences
in the nature of Moran interactions between central CA and southern CA (Results), specifically
differences between the two regions in the lag of nitrate effects on kelp populations. 
If climate change modifies environmental effects on populations in a related way it would 
be expected to produce similarly large changes in Moran interactions. 
Climate change may alter 
lags and delays associated with environmental effects on populations in at least two ways:
1) by altering species phenology; and 2) by increasing or decreasing growth rates and thereby
decreasing or increasing delays. Though it is too early to conclude that
effects on Moran interactions are among the most important impacts of climate change, we feel the mechanisms
outlined above are sufficient to warrant further investigation.

@@

%*Revisit the kendall paper, briefly. Talk about possible three-way interactions.
We have focussed on interactions between two Moran drivers, but synchrony in most systems
may be a phenomenon with multiple (more than two) interacting causes.
\cite{Kendall2000} considered interactions between dispersal and a Moran mechanism of synchrony. Their research
questions were therefore distinct from ours, but combining their viewpoints and ours may lead to 
future work about interactions between dispersal and more than one Moran driver of synchrony. 
Although we considered only two Moran drivers in 
our theory and examples, essentially all population systems are influenced by multiple environmental drivers,
and environmental drivers very commonly are associated with large scale climatic phenomena
such as ENSO, and hence are associated with each other. Thus it may be quite common for synchrony to simultaneously be caused by dispersal and
multiple distinct Moran effects, and these influences may interact in multifarious combinations. 
It may be necessary in future
work to consider interactions between dispersal and two Moran drivers. 
It may be useful to consider cases for which multiple related Moran drivers all interact.
Dispersal can readily be added to our modelling framework:
\cite{Desharnais2018} performed a spectral analysis on a model similar to ours
which included dispersal.
Future work should consider whether and when scenarios of multi-driver interactions between causes of synchrony 
can lead to synchrony patterns which differ fundamentally from what one would expect from one
or two mechanisms. Potential interactions increase as the square of the number of 
drivers, so interactions seem likely to become more important as our viewpoint of synchrony expands.

>>>>

%length stats
%Intro through Discussion: 7719 as of 2023 02 08 (limit is 7500)
%7 figures, 1 table (limit is 10)
%so I think I'm OK - slightly over on the words, but with 2 fewer displays

@@

\affil[a]{Department of Ecology and Evolutionary Biology and Center for Ecological Research, University of Kansas}
\affil[b]{Department of Environmental Sciences, University of Virginia}
\affil[c]{Department of Geography, University of California, Los Angeles}
\affil[d]{Marine Biological Association of the United Kingdom}
\affil[e]{Center for Watershed Sciences, University of California, Davis}
\affil[f]{Department of Applied Ocean Physics and Engineering, Woods Hole Oceanographic Institution}
\affil[g]{Earth Research Institute, University of California, Santa Barbara}
\affil[*]{Corresponding author, reuman@ku.edu}

@@

%Intro to Moran effects and history of the early years where people had trouble identifying them
%***DAN: new review para for ecography
Correlations between weather time series measured in different locations can 
induce synchrony between populations in those locations if the weather variables influence
population processes. This mechanism, called the Moran effect, is now known to be one of the most
important causes of synchrony. But the mechanism was originally proposed 
theoretically \citep{Moran1953}, and decades passed during which it was 
considered difficult to distinguish this potential cause of synchrony from others in field systems.
Synchrony has long been thought to have three causes: dispersal, and trophic interactions with a synchronous or
mobile species, in addition to the Moran effect \citep{Liebhold2004}. However, using common past statistical approaches
which focussed on declines in population correlations with distance, patterns of synchrony produced
by each of these mechanisms are quite similar \citep{Ranta1999,Abbott2007,Liebhold2004,Walter2017}; so examination
of such patterns provides little or no traction for inferring the causes of synchrony. 
Early papers demonstrating the Moran effect mechanism resorted to special cases where dispersal was 
impossible and predators were absent \citep{Grenfell1998,Tedesco2004}. Controlled experiments also confirmed that all
three theorized mechanisms could be involved in synchrony [e.g., \citealp{Vasseur2009}]. Nevertheless, the problem of inferring
specific mechanisms of synchrony in field systems was considered a challenge 
until recently.

%Recent statistical explorations of Moran effects
%***DAN: new review para for ecography
Recent research has provided new statistical viewpoints which have, when sufficient data are available, 
essentially solved the problem of inference
of the causes of synchrony, and the research has revealed the broad importance of Moran effects. 
%geographic approaches
Approaches based on formal statistical comparisons between population synchrony maps and maps of 
environmental correlations produced inferences that precipitation and temperature Moran effects are important
causes of synchrony of terrestrial \citep{Defriez2017b} and marine \citep{Defriez2017} primary productivity. 
Another geographic approach, based on multiple regression of distance matrices (MRM), was used in an early paper
to infer that a precipitation Moran effect is a cause of synchrony for the spongy moth \citep{Haynes2013},
an invasive forest pest whose name was recently changed from ``gypsy moth.'' 
Geographic approaches to identifying causes of synchrony, often MRM approaches 
specifically \citep{Koenig2017,Walter2020,Bogdziewicz2021}, 
have become widespread, and the geography of synchrony has become a mainstream area of study \citep{Walter2017}.
The approach of \cite{Gouveia2016}, called ``noncentered local indicators of spatial association'' (ncLISA), is another 
geographic approach; as is the ``fluvial synchrogram'' of \cite{Larsen2021}, a tool for studying synchrony in 
river networks.
MRM approaches have also been used to identify dispersal as a cause of synchrony \citep{Anderson2018},
sometimes combined with genetic methods \citep{Haynes2022}.
%Time and timescale methods
Another class of methods exploit the time and timescale structure of synchrony.
Using a method that examined the match between temporal variations in population and environmental synchrony, 
\cite{Allstadt2015} confirmed the importance of a precipitation Moran effect for synchrony of spongy moth.
A suite of wavelet and related Fourier techniques have been developed that can comprehensively describe the time and timescale
structure of synchrony, identify Moran drivers of synchrony opperating in distinct timescale bands, and apportion fractions
of observed synchrony to respective Moran drivers \citep{Sheppard2016,Sheppard2017,Desharnais2018,Sheppard2019,wsyn,Anderson2021}.
The methods have been used, for instance, to provide one of the first clear demonstrations that changes in 
Moran effects are another of the impacts of climatic changes \citep{Sheppard2016,Sheppard2019}; and to illustrate the importance of the
timescale structure of environmental covariation for altering Moran effects \citep{Desharnais2018}.
These methods have since been used to identify Moran effects in several additional systems
(e.g., \citealp{Anderson2019,Walter2019,Walter2020,GarciaCarreras2022,CastoraniSheppardKelp}).

%Intro to interactions
The techniques reviewed above have made it possible for several recent papers to identify cases in which two or more
distinct Moran drivers opperate simultaneously on the same populations 
(e.g., \citealp{Defriez2017b,Haynes2019,Walter2019,Walter2020,Anderson2021}).
In fact, two recent papers have documented that Moran effects of distinct 
environmental drivers can interact, either synergistically or antagonistically,
so that total population synchrony can be either greater than or less than what synchrony would be if the distinct Moran 
drivers operated independently \citep{Sheppard2019,CastoraniSheppardKelp}. 
\cite{Sheppard2019} showed that long-timescale ($>4$y period) synchrony in a chlorophyll density index 
in the seas around the United Kingdom is substantially augmented by interactions between environmental 
drivers of synchrony and drivers linked with consumption by a copepod consumer.
\cite{CastoraniSheppardKelp} showed that nutrient dynamics and wave disturbance, two Moran drivers
of synchrony in giant kelp populations on the California (CA) coast, interact either synergistically 
or destructively, depending on which portion of the CA coast is examined and on 
the timescale of analysis.

%main purpose
The main purpose of this study is to establish a general theory of the mechanism
of interacting Moran effects and to use examples and applications of the theory to provide 
ecological intuition for how the mechanism of interacting Moran effects works. Our goals are
distinct from those of \cite{Kendall2000}, who examined interactions between Moran and dispersal
causes of synchrony.

@@

%Direct Moran effects
Direct Moran effects for kelp and how they manifest in our theory are fairly straightforward. 
Nitrates are henceforth identified with $\epsilon^{(1)}$ and waves with $\epsilon^{(2)}$. 
Both nitrates and waves fluctuate seasonally in CA \citep{Schiel2015}. 
Thus nitrate and wave 
synchrony had a strong annual component (Fig. \ref{fig:noise}a,b,j,k for the CCA1 and SB regions,
Fig. \ref{SM-fig:directMoran_SM_lag4} for CCA2), which produced 
some of the annual synchrony observed in kelp 
(Fig. \ref{fig:SummaryKelpFig}a,d, green and blue lines).
Nitrates and waves are also synchronous on long timescales $>8$y  
(Fig. \ref{fig:noise}g,h,p,q for the CCA1 and SB regions, Fig. \ref{SM-fig:directMoran_SM_lag4} 
for CCA2), possibly due to the North Pacific Gyre Oscillation 
\citep{CastoraniSheppardKelp, DiLorenzo2008}. The long-timescale synchrony
in nitrates and waves produced some of the long-timescale
synchrony in kelp (Fig. \ref{fig:SummaryKelpFig}c,f, green and blue lines).
Kelp synchrony was stronger in CCA1 than in SB 
(Fig. \ref{fig:SummaryKelpFig}a-c v. d-f, black lines) in part because the synchrony of nitrates, and of 
waves, was more pronounced in CCA1 than in SB 
(Fig. \ref{fig:noise}), and also because waves
had a stronger influence on kelp in central California than in SB:
regression coefficients determining $f_{\mcP^{(1)}}$, $f_{\mcP^{(2)}}$
and $f_{\mcB}$ are in Table \ref{SM-tab:regcoefs}.

@@

%*DONE Climate change
Climate change has the potential to influence interactions between Moran effects in two specific ways
which can be illuminated by our theory, and this potential should be investigated in
future work. Examining the third term of (\ref{rhoww_T1to3}), climate change could
alter interaction effects if it: 1) alters the term $\rho_{\epsilon^{(1)} \epsilon^{(2)}}$
quantifying cross synchrony between Moran drivers; or 2) alters one of the terms
$f_{\mcP^{(1)}}$ or $f_{\mcP^{(2)}}$ specifying the nature of the influence of the 
environmental variables $\epsilon^{(1)}$ and $\epsilon^{(2)}$, respectively, on populations. As advocated above,
the term $\rho_{\epsilon^{(1)} \epsilon^{(2)}}$ should be systematically computed in future work, for a variety 
of environmental variables, to assess whether interactions between Moran effects are likely to
be general. As part of that process, the potential for changes in $\rho_{\epsilon^{(1)} \epsilon^{(2)}}$
could also be assessed, by using either time-windowed versions or wavelet adaptations 
of this statistic, applied to either long-term climate records or future climate scenarios. 
Differences in $f_{\mcP^{(1)}}$ were responsible for differences
in the nature of Moran interactions between central CA and southern CA (Results), specifically
differences between the two regions in the lag of nitrate effects on kelp populations. 
If climate change modifies environmental effects on populations in a related way it would 
be expected to produce similarly large changes in Moran interactions. 
Climate change may alter 
lags and delays associated with environmental effects on populations in at least two ways:
1) by altering species phenology; and 2) by increasing or decreasing growth rates and thereby
decreasing or increasing delays. Though it is too early to conclude that
effects on Moran interactions are among the most important impacts of climate change, we feel the mechanisms
outlined above are sufficient to warrant further investigation.

@@

%*Revisit the kendall paper, briefly. Talk about possible three-way interactions.
%***DAN: new review para for ecography
We have focussed on interactions between two Moran drivers, but synchrony in most systems
may be a phenomenon with multiple (more than two) interacting causes.
\cite{Kendall2000} considered interactions between dispersal and a Moran mechanism of synchrony. Their research
questions were therefore distinct from ours, but combining their viewpoints and ours may lead to 
future work about interactions between dispersal and more than one Moran driver of synchrony. 
Although we considered only two Moran drivers in 
our theory and examples, essentially all population systems are influenced by multiple environmental drivers,
and environmental drivers very commonly are associated with large scale climatic phenomena
such as ENSO, and hence are associated with each other. Thus it may be quite common for synchrony to simultaneously be caused by dispersal and
multiple distinct Moran effects, and these influences may interact in multifarious combinations. 
It may be necessary in future
work to consider interactions between dispersal and two Moran drivers. 
It may be useful to consider cases for which multiple related Moran drivers all interact.
Dispersal can readily be added to our modelling framework:
\cite{Desharnais2018} performed a spectral analysis on a model similar to ours
which included dispersal.
Future work should consider whether and when scenarios of multi-driver interactions between causes of synchrony 
can lead to synchrony patterns which differ fundamentally from what one would expect from one
or two mechanisms. Potential interactions increase as the square of the number of 
drivers, so interactions seem likely to become more important as our viewpoint of synchrony expands.
